---
title: "Titanic Data Analysis (Interpretation and Prediction)"
author: "Sandipan Roy"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
      code_download: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

The sinking of the Titanic is one of the most infamous shipwrecks in history.


On April 15, 1912, during her maiden voyage, the widely considered “unsinkable” RMS Titanic sank after colliding with an iceberg. Unfortunately, there weren’t enough lifeboats for everyone onboard, resulting in the death of 1502 out of 2224 passengers and crew.


While there was some element of luck involved in surviving, it seems some groups of people were more likely to survive than others.

In this problem, you are given two similar datasets that include passenger information like name, age, gender, socio-economic class, etc. One dataset is titled ''train.csv'' and the other is titled ''test.csv''.

''train.csv'' will contain the details of a subset of the passengers on board (891 to be exact) and importantly, will reveal whether they survived or not, also known as the “ground truth”. The ''test.csv'' dataset contains similar information but does not disclose the ''ground truth'' for each passenger. 

Using the patterns you find in the train.csv data, predict whether the other 418 passengers on board (found in ''test.csv'') survived.

\underline{Variable Definition}

* **survival**: A binary variable indicating whether the individual survived or not. 0=No, 1=Yes
* **pclass**: Ticket class which is a proxy for socio economic status. 1=1st, 2=2nd, 3=3rd
* **sex**: Male or Female
* **Age**: Age in years. Note that age is fractional if less than 1. If the age is estimated, is it in the form of xx.5
* **sibsp**: number of siblings / spouses aboard the Titanic. The dataset defines family relations in this way- Sibling = brother, sister, stepbrother, stepsister. Spouse = husband, wife (mistresses and fiancés were ignored)
* **parch**: number of parents / children aboard the Titanic. The dataset defines family relations in this way- Parent = mother, father. Child = daughter, son, stepdaughter, stepson. Some children travelled only with a nanny, therefore parch=0 for them.
* **ticket**: Ticket number
* **fare**: Passenger fare
* **cabin**: Cabin number
* **embarked**: Port of Embarkation. C = Cherbourg, Q = Queenstown, S = Southampton

1. Read the data and present a summary of the dataset. Are there any missing values? Comment on the distribution of sex according to the ticket class.

```{r}
test <- read.csv('test.csv')
train <- read.csv('train.csv')
```

2. Use a suitable plot to comment on the male and the female survival according to the ticket class.

```{r}
train %>% 
  group_by(Pclass, Sex) %>% 
  summarise(Survived = mean(Survived)) %>% 
  ggplot2::ggplot() +
  ggplot2::geom_bar(aes(x = Sex, y = Survived), stat = 'identity') +
  facet_wrap(~ Pclass)
```


3. **(H)** Fit an appropriate GLM (without interaction) with sex and ticket class as predictors and survival as response variable. Interpret the coefficients in the fitted model.

4. **(H)** Fit an appropriate GLM (with interaction) with sex and ticket class as predictors and survival as response variable. Interpret the coefficients in the fitted model. How does the female survival changes in this model compared to the model in question 3? 

5. **(H)** Predict the survival of the passengers given in the test dataset. Also, provide the prediction error and the cross validation error in the training dataset.